---
title: "Tidy data"
author: "Ashwin Malshe"
date: "06/01/2021"
output:
  pdf_document: default
  html_document:
    theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(printr) # This package prints help in the rmarkdown document after knitting
```

For this exercise we will use `mpg` dataset, which is bundled with `ggplot2`. You don't have to explicitly call `ggplot2` into this R session because we loaded it already using `tidyverse`.

## `mpg` dataset

First, let's understand what is inside this dataset. A `?` followed by the name of the object/function/package will lead to opening up the help menu on right.

```{r eval=FALSE}
?mpg # OR help(mpg)
```

### Exercise 1: Explore

In this exercise, we will explore the data set.

`summary()` will output simple summary for all the variables in the data set. This is a Base R function.

```{r}
summary(mpg)
```

Let's look at the frequency of character variables using `table()` function

```{r}
table(mpg$manufacturer)
```

`table()` outputs a vector of frequencies. What if you wanted a dataframe in return, instead? We will use `count()` function from `dplyr`

```{r}
count(mpg, manufacturer, model, year)

```

It will work on numeric variables as well.

```{r}
count(mpg, year)
```

#### Printing specific statistics

Use these functions to calculate summary statistics for `hwy` variable.

```{r eval = FALSE}
mean()
median()
sd()
min()
max()
```

```{r}
mean(mpg$cty)
```

```{r}
mad(mpg$cty)
```

```{r}
sd(mpg$cty)
```

### Create a dataframe with the summary statistics of `hwy`

*Here also introduce `pipe` operator `%>%`*

```{r}
sum(5, 7)
```

```{r}
as.character(sum(5, 7))
```

```{r}
paste("My number is", as.character(sum(5, 7)))
```

```{r}
a <- sum(5, 7)
b <- as.character(a)
paste("My number is", b)
```

```{r}
5 %>% 
  sum(7)
```

```{r}

  as.character(5 %>% sum(7))
```

```{r}
5 %>% 
  sum(7) %>% 
  as.character() %>% 
  paste("My number is")
  
```

```{r}
5 %>% 
  sum(7) %>% 
  as.character() %>% 
  paste("My number is", .)
  
```

```{r}
mpg_sum1 <- summarize(mpg,
                      hwy_mean = mean(hwy),
                      hwy_med = median(hwy),
                      hwy_sd = sd(hwy))
```

```{r}
mpg_sum1 <- mpg %>% 
  summarize(hwy_mean = mean(hwy),
         hwy_med = median(hwy),
         hwy_sd = sd(hwy))
```

```{r}
print(mpg_sum1)
```

`dplyr` has a new `across()` function, that allows us to perform multiple operations on multiple variables in one shot.

Imagine, you want the mean and standard deviations of `hwy` and `cty`. Here is how to get them in one shot.

```{r}
mpg %>% 
  summarize(across(.cols = c(hwy, cty), 
                   .fns = list(mean = mean, sd = sd)))
```



### Grouping and summarizing

A common operation in data prep is to group datasets by a variable and then summarize the variable.

For instance, above we got summary statistics for all the full data set. What if we wanted it separately for 1999 and 2008?

```{r}
mpg_sum2 <- mpg %>% 
  group_by(year) %>% 
    summarize(across(hwy, list(mean = mean, median = median, sd = sd))) %>% 
  ungroup()
```

```{r}
print(mpg_sum2)
```

```{r}
mpg %>% 
  group_by(manufacturer, year) %>% 
    summarize(across(hwy, list(mean = mean, median = median, sd = sd))) %>% 
  ungroup()
```

*Experimental feature using `.groups`*

```{r}
mpg_sum3 <- mpg %>% 
  group_by(year) %>% 
    summarize(across(hwy, list(mean = mean, median = median, sd = sd)),
           .groups = "drop") 
```

```{r}
print(mpg_sum3)
```

## Merging data frames

The simplest task is to attach data frames by columns or rows without paying any consideration for a common variable to match on. For this, R has two functions:

### `rbind`

If we want to put data frames on top of each other, we can use `rbind()`

For `rbind()` to work, the two data frames should have:

1.  The same number of columns
2.  The same column names
3.  The same `class` of columns

The easiest demo for this can be created by splitting an already existing data frame. Let's separate out `mpg` by year and then merge the two resulting data frames back

```{r}
mpg_1999 <- mpg %>%  filter(year == 1999)
mpg_2008 <- mpg %>%  filter(year == 2008)
```

```{r}
mpg_rows <- rbind(mpg_1999, mpg_2008)
```

```{r}
summary(mpg_rows)
```

### `cbind`

If you want to merge two data frames sideways, you can use `cbind()`

For it to work,

1.  The two data frames should have the same number of rows.

Now, let's first split `mpg` along the columns. First data frame will have first 5 columns and the second data frame will hold the remaining columns

```{r}
names(mpg)
```

```{r}

mpg_left <- mpg %>% 
  select(1:5) %>% 
  mutate(id = row_number()) # you will know what this does in a moment

mpg_right <- mpg %>% 
  select(-c(1:5)) %>% 
  mutate(id = row_number())

```

```{r}
mpg_cols <- cbind(mpg_left, mpg_right)
```

```{r}
summary(mpg_cols)
```

### `*_join()` functions from `dplyr`

Commonly, you need to merge data sets sideways by adding columns. However, there are two common issues:

1.  The two data sets may not have the same number of rows
2.  Even if the two data sets have the same number of rows, there is no guarantee that you are matching the rows correctly.

Let's understand the second issue first. Sort `mpg_right` by `cty`

```{r}
mpg_right2 <- mpg_right %>% 
  arrange(cty)

```

Now if you use `cbind()` on `mpg_left` and `mpg_right2` you won't get back the same original data frame. This is because, we have mismatched the observations.

How can we resolve this issue?

```{r}
mpg_cols <- inner_join(mpg_left, mpg_right2, by = "id")
```

### Merge summary back to the original data frame

The common key helps even when we don't have the same number of observations in the two data frames

```{r}
print(mpg_sum2)
```

```{r}
names(mpg)
```

```{r}
mpg2 <- mpg %>% filter(year == 1999) %>% 
  right_join(mpg_sum2, 
             by = "year")
```

```{r}
dim(mpg2)
```
