---
title: "Functions and Loops"
author: "Ashwin Malshé"
date: "10/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Functions wrap longer procedures in one single command

Next we will create a function called forever21.

```{r}
forever21 <- function(age) {
  if (age <= 21) {
    return(age)
  } else {
    return(21)
  }
}
```


```{r}
forever21(50)
forever21(15)
```

Write a function to add two numbers

```{r}
mySum <- function(a, b) {
  return(a + b)
}

mySum(5, 10)
mySum(255, 255)
```

Write a function to concatenate two character variables
For example, you input "Hello" and "world!" separately and you get "Hello world!"
Recall that we modified this funciton in the class to accommodate output with multiple strings

```{r}
myWord <- function(a, b) {
  w1 <- paste(a, b, sep = "|")
  w2 <- paste(a, b, sep = "-")
  w3 <- paste(a, b, sep = "***")
  return(c(w1, w2, w3))
}
myWord("Hello", "world!")
```

Write a function to winsorize a vector at a given percentage

```{r}
winsor <- function(vec, w = 0.95) {
  l1 <- (1 - w) / 2
  l2 <- 1 - l1
  q <- quantile(vec, c(l1, l2), type = 1)
  vec1 <- ifelse(vec < q[1], q[1], vec)
  vec2 <- ifelse(vec1 > q[2], q[2], vec1)
  return(vec2)
}
```



```{r}
winsor(vec = c(1:100), w = 0.9)
```



## Writing functions with default values

Write a function for area of a circle with radius r. The default value should be for radius r = 1

```{r}
cirArea <- function(r = 1) pi * r^2
```

get area of a circle with radius = 5. Verify the result

```{r}
# Answer using the function
cirArea(5)
# Answer using the raw formula
pi * 5^2

identical(cirArea(5), pi * 5^2)
```


## Saving and retrieving functions
Copy the function and paste it in an R script. Save the file on your hard disk and make note of the path to that file.

When you want to load the function, use 

`source("path name to the function R file")`


## Looping in R

```{r}
for (i in 1:10) {
  print(paste("My number is", i))
}
```

Also check out `while` command in R. It can be used when the condition is not easy or possible to write in a `for`. For example, you may want to keep running the loop until specific time (e.g., 5 pm) or until you get a specific number of observations from an operation (e.g., download 5,000 tweets)

```{r}
i <- 1
while(i <= 10) {
  print(paste("My number is", i))
  i <- i + 1
}
```








