---
title: "tidyverse basics"
author: "Ashwin Malshe"
date: "06/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

In R you can install packages by typing this in the console:

`install.packages("NAME of the PACKAGE")`

You will need an active internet connection because R will download the package content and then install it. This is a one-time task so never include installation statements in your code.

Let's create a data frame to work with.

```{r}
d1 <- data.frame(name = c("John", "Mary", "Toby", "Maya", "Don"),
                 age = c(20, 21, 45, 78, 90),
                 male = c(TRUE, FALSE, TRUE, FALSE, TRUE))

print(d1)
```

### Exercise 1

Use the data frame `d1` and keep all the rows where `age` is <= 50.


For this task, we need to use `filter()` from `dplyr`

```{r}

d2 <- filter(d1,
             age <= 50)

d2
```

### Exercise 2

Use the data frame `d1` and remove all the rows where `age` is > 50 and `male` is TRUE

```{r}
d3 <- filter(d1,
             age > 50 & male == TRUE)

d3
```

Alternatively,

```{r}
d3 <- filter(d1,
             age <= 50 | male == FALSE)

d3
```

### Exercise 3

From `d1` keep only the first two columns labeled `name` and `age`

Here we use `select()` function:

```{r}
select(d1, name, age)
```

### Exercise 4

Extract all the columns with names starting with letter "n".

```{r}
select(d1, starts_with("n"))
```

### Exercise 5

Extract all the numeric columns


```{r}
select(d1, is.numeric)
```

### Exercise 6

Rename `name` as `first_name` and `age` as `age_in_years`

```{r}
rename(d1,
       first_name = name,
       age_in_years = age)
```

